A recipe for extremely reproducible enrichment analysis (and why we need it)

### 2023-12-11 manuscript/manuscript.Rmd
manuscript updates

In addition to shared code, data and containerisation, there are a few other ways researchers can make
their work more reproducible.
Thorough documentation makes the reproducibility process a great deal easier and more
accessible, especially to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using R Markdown or Jupyter notebooks provides optimal documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Code and data can be linked. This means that the code "knows" the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one or a few
commands, and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum, and the *five pillars*
framework has been described as a way of achieving extreme reproducibility @Ziemann2023-dg.
Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
over a decade.

@@

This irreproducibility might be expected given that substantial methodological flaws and reporting
deficiencies were present in 85% of enrichment analyses from which this set of 20 studies were
drawn @Wijesooriya2022-li.

@@

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at risk
of becoming unavailable due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma, and allows users to download the computer script to facilitate
future non-interactive reproducibility @Powell2019-lv.
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions[@Ge2020-qe], which aids future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"
approach detailed here.

>>>>

In addition to shared code, data and containerisation, there are a few other ways researchers can make
their work more reproducible.
Thorough documentation makes the reproducibility process a great deal easier and more
accessible, especially to novice bioinformaticians.
Version control keeps track of code and documentation changes over time and manages the edits made by
multiple contributors.
Literate programming using R Markdown or Jupyter notebooks enables thorough documentation by embedding
free text explanations together with the code and outputs such as charts and tables.
Code and data can be linked. This means that the code "knows" the location of the data,
avoiding any "file not found" errors @Peng2011-ia.
Lastly, reanalysis can be made as simple as possible by creating a master script that executes analysis
for each part sequentially, making the entire manuscript reproducible by executing just one or a few
commands, and without any manual steps @Lasser2020-mm.
Taken together, computational reproducibility is a spectrum, and the *five pillars*
framework has been described as a way of achieving extreme reproducibility @Ziemann2023-dg.
Currently, bioinformatics work at this gold standard is rare, despite the availability of such tools for
over a decade.

@@

This irreproducibility might be expected given that substantial methodological flaws and reporting
deficiencies were present in 85% of enrichment analyses from which this set of 20 studies were
drawn @Wijesooriya2022-li.
Authors can prevent these reporting decifiencies providing a checklist of methodological details similar
to ours (**Table 2**).
Checklists like these are a useful resource as they are simple for journals to enforce and reviewers to check.

@@

With this, we must reconsider whether web-based tools are valid for scientific publications if they are at
risk of becoming unavailable wthout notice due to link decay.
Some progress is being made to make web-based point-and-click tools more reproducible.
The Galaxy, GenePattern and RNAlysis projects offer graphical tools to create, edit and import
workflows that enable reproducibility [@Kuehn2008-su; @Goecks2010-lq; @Teichman2023-fx].
The Degust differential expression suite is a graphical interface to popular
expression tools like edgeR and limma, and allows users to download the computer script to facilitate
future non-interactive reproducibility @Powell2019-lv.
The ShinyGO web-based enrichment tool is a satisfactory GUI-based alternative to DAVID as it provides Docker image
downloads for all past versions[@Ge2020-qe], facilitating preservation and aiding future reproducibility.
Still, such tools leave documentation and reproducibility gaps which are resolved with the "extremely reproducible"
approach detailed here.
